---
title: Syllabus
slug: syllabus
citeproc: false
bibliography: ../static/bib/references.bib
csl: ../static/bib/chicago-syllabus-no-bib.csl
---

<!-- \bibliography{../static/bib/references.bib} So that RefTeX knows about the bibliography -->

```{r load-libraries, message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
library(tibble)
library(pander)
```

# Objectives

This bootcamp will do two things. First, it will teach you how to use modern tools to create insightful, beautiful, reproducible visualizations of data. Second, you will learn about the theory and practice of efforts to visualize data. We will think about different ways of looking at data, where data comes from, and the implications of chosing to represent it in different ways.

By the end of the workshop series you will

- Understand the basic principles behind effective data visualization.
- Have a practical sense for why some graphs and figures work well,
  while others may fail to inform or actively mislead.
- Know how to clean, manipulate, and save datasets in R using dplyr and tidyr.
- Know how to create a wide range of plots in R using ggplot2.
- Have an understanding of some issues surrounding the collection and 
  representation of data.
- Have exposure to some basic data science methods in R.

# Recommended Readings

I strongly recommend getting ahold of the following books. The online or draft versions are fine, don't feel like you must purchase them:

- [@healy:2019] [Draft version free online; print version at [Amazon](https://amzn.to/2vfAixM) or other bookshops.]
- [@wickhamgrolemund:2017] [Free online; print version at [Amazon](https://www.amazon.com/R-Data-Science-Hadley-Wickham/dp/1491910399/) or other bookshops.]
- [@wilke:2019] [Draft version free online; print version at  [Amazon](https://www.amazon.com/Fundamentals-Data-Visualization-Informative-Compelling/dp/1492031089) or other bookshops.]

# Software

We will do all of our visualization work in this class using the programming language [R](https://cran.r-project.org/). We will use [RStudio](https://www.rstudio.com/) to manage our code and projects. R and RStudio are widely used tools for data analysis in academia and industry.

*Come to the first session on February 11 if you wish to have help completing the following.*

You will need to install some software first. You can consult the [orientation material](https://jasonheppler.org/courses/bootcampr.2020/reading/01-reading/), but in brief here's what to do: 

1. Get the most recent version of R. R is free and available for Windows, Mac, and Linux operating systems. Download^[`cloud.r-project.org`] the version of R compatible with your operating system. If you are running Windows or MacOS, you should choose one of the *precompiled binary* distributions (i.e., ready-to-run applications) linked at the top of the R Project's webpage. 
2. Once R is installed, download and install R Studio.^[`rstudio.com`] R Studio is an "Integrated Development Environment", or IDE. This means it is a front-end for R that makes it much easier to work with. R Studio is also free, and available for Windows, Mac, and Linux platforms.
3. Install^[`tidyverse.org`] the tidyverse library and several other add-on packages for R. These libraries provide useful functionality that we will take advantage of throughout the book. You can learn more about the tidyverse's family of packages at its website.

   To install the tidyverse and some additional useful packages, make sure you have an Internet connection and then launch R Studio. Type the following lines of code at R's command prompt, located in the window named "Console", and hit return. In the code below, the `<-` arrow is made up of two keystrokes, first `<` and then the short dash or minus symbol, `-`.

```{r, echo=TRUE, eval=FALSE, tidy = FALSE}
my_packages <- c("tidyverse", "broom", "gapminder", "GGally", "ggraph", 
                 "ggrepel", "ggridges", "gridExtra", "here", 
                 "maps", "mapproj", "mapdata", "rlang", "scales", 
                 "sp", "usethis", "devtools")

install.packages(my_packages, repos = "http://cran.rstudio.com")
```

R Studio should then download and install these packages for you. It may take a little while to download everything.

You may, alternatively, choose to use [rstudio.cloud](https://rstudio.cloud), which is currently free for use but I must also warn you that it's alpha software and may not remain stable. RStudio Cloud is great for running RStudio in the browser, thus eliminating the need for you to install R and related software on your own machine. 

# Schedule

`r blogdown::shortcode("topiclist")`

As the weeks go by, consult the [Schedule Page](/schedule/) for more information on topics, worksheets, and readings. The schedule may change as we go. Links to readings and other materials will be posted on that page. 

# Policies

- Attendance to all workshops is not required, but be aware that I won't have time during the workshops to help troubleshoot technical issues. If you're running into issues during a workshop, you'll likely need to figure out a solution on your own or schedule a time to meet with me afterwards. If you're running into issues outside of the workshop, please set up an appointment with me and we can troubleshoot together.
- Please try and complete the readings in advance of the workshop. 
- Feel free to bring a laptop to the workshop and use it for note-taking and in-class work. 
